---
title: "Bender et al (2024) -- EMBO"
subtitle: shRNA screen analysis
author:
- name: Alexander Bender
  affiliation: Institute of Molecular Tumor Biology, Muenster/Germany
date: "`r paste('Compiled:', format(Sys.time(), '%d-%b-%Y'))`"
output:
  rmdformats::readthedown:
    code_folding: show
    keep_md: false
    highlight: tango
    toc_float:
      collapsed: false
editor_options: 
  markdown: 
    wrap: 200
params:
  save_final: true
---

<style>
body {
text-align: justify}
</style>

# Setup

Define root directory that contains the folder with source data. Run script that
loads packages and define document-specific variables.

```{r setup}

# Inside this Docker container we mount the directory with all the source data as "/projectdir/"
rootdir <- "/projectdir/"
source(paste0(rootdir, "/runStartup.R"))
```

# shRNA screen

# Infection bias

Don't be surprised. What in the paper we call URE-AML is here called B22. URE is Hox-URE, and WT is Hox-WT.

```{r shrna_inf_bias}

# The raw counts based on the bowtie2 alignment
raw_counts_shrna <- read.delim(
  file = paste0(rootdir, "/source_data/GSE250630_shrna_screen_rawCounts.tsv.gz"),
  header = TRUE, row.names = "shrna"
) %>% dplyr::select(-gene)

# sequences <- openxlsx::read.xlsx(paste0(rootdir, "/source_data/shrna_screen_sequences.xlsx"), check.names = FALSE)

# Make pairwise comparisons to explore data -- normalize to NTCs
y <- edgeR::DGEList(counts = raw_counts_shrna, group = gsub("_rep.*", "", colnames(raw_counts_shrna)))

y_ntc <- y[grepl("NTC", rownames(y)), ]
y$samples$norm.factors <- edgeR::calcNormFactors(y_ntc)$samples$norm.factors
rm(y_ntc)

design <- stats::model.matrix(~ 0 + group, y$samples)
colnames(design) <- gsub("group", "", colnames(design))

v <- limma::voom(y, design = design)

contrasts_voom <- limma::makeContrasts(
  WT_Input_vs_URE_Input = HoxWT_Input - HoxURE_Input,
  WT_Input_vs_B22_Input = HoxWT_Input - B22_Input,
  URE_Input_vs_B22_Input = HoxURE_Input - B22_Input,
  WT_Final_vs_WT_Input = HoxWT_Final - HoxWT_Input,
  URE_Final_vs_URE_Input = HoxURE_Final - HoxURE_Input,
  B22_Final_vs_B22_Input = B22_Final - B22_Input,
  levels = design
)

res_voom <- sapply(colnames(contrasts_voom), function(x) {
  fit <- limma::lmFit(v, design)
  con <- contrasts_voom[, x]
  cf <- limma::contrasts.fit(fit, contrasts = con)
  cf <- limma::eBayes(cf)

  limma::topTable(cf, number = Inf) %>%
    dplyr::mutate(shRNA = rownames(.)) %>%
    dplyr::rename(FDR = adj.P.Val, baseMean = AveExpr)
}, simplify = FALSE)

# Visualize the comparison of (end of screen) vs (input screen)
use_barcode_levels <- c("killing", "NTC", "signature3")
use_barcode_alphas <- c(1, 1, 0.35)
use_shRNA_pointsize <- 1.5

elements <- list()

# this stores the elements of Figure7B and C in the list figure7_elements
names(res_voom) <- gsub("WT_", "Hox-WT_", gsub("URE_", "Hox-URE_", gsub("B22_", "URE-AML_", names(res_voom))))
for (x in names(res_voom)) {
  tmp_ylims <- unlist(lapply(res_voom, function(x) {
    c(min(x$logFC), max(x$logFC))
  }))

  elements[[paste0("voom_", x)]] <-
    res_voom[[x]] %>%
    mutate(
      controls = factor(case_when(
        grepl("NTC", shRNA) ~ "NTC",
        grepl("Psma|Polr2b|Rpl30", shRNA) ~ "killing",
        TRUE ~ "signature3"
      ),
      levels = use_barcode_levels
      ),
      signif = factor(case_when(
        logFC > 0 & FDR < 0.05 ~ "up",
        logFC < 0 & FDR < 0.05 ~ "down",
        TRUE ~ "nonsig"
      ),
      levels = c("up", "down", "nonsig")
      )
    ) %>%
    arrange(controls, decreasing = TRUE) %>%
    ggplot(aes(x = baseMean, y = logFC, color = controls)) +
    geom_point(aes(size = controls, alpha = controls)) +
    scale_color_manual(values = c("firebrick", "darkblue", "grey60")) +
    scale_size_manual(values = c(use_shRNA_pointsize, use_shRNA_pointsize, use_shRNA_pointsize / 2)) +
    scale_alpha_manual(values = use_barcode_alphas) +
    ggtitle(gsub("_Input", "", gsub("_Input_vs_", " vs ", x))) +
    geom_hline(yintercept = 0, lty = 2, size = .25) +
    ylim(c(floor(min(tmp_ylims)), ceiling(max(tmp_ylims)))) +
    theme(
      legend.position = "top",
      legend.title = element_blank(),
      legend.justification = "left",
      legend.margin = margin(0, -.25, 0, 0)
    ) +
    guides(size = guide_legend(override.aes = list(size = 3, alpha = 3))) +
    theme(legend.text = element_text(size = 10))

  # this is simply to ensure that colored dots (blue/red) are on top of the grey ones
  tmp <- elements[[paste0("voom_", x)]]
  tmp$data <- tmp$data %>%
    mutate(ID = seq(1, length(controls))) %>%
    arrange(-ID)
  elements[[paste0("voom_", x)]] <- tmp
  rm(tmp)
}

# Show that there is no infection bias between celllines
Appendix_Figure_S3B <-
  ((elements$`voom_Hox-WT_Input_vs_Hox-URE_Input` + ylim(c(-2, 2)) + ggtitle("")) |
    (elements$`voom_Hox-URE_Input_vs_URE-AML_Input` + ylim(c(-2, 2)) + ggtitle("")) |
    (elements$`voom_Hox-WT_Input_vs_URE-AML_Input` + ylim(c(-2, 2)) + ggtitle(""))) +
    plot_layout(guides = "collect") & theme(legend.position = "top")

Appendix_Figure_S3B

# Not in the paper but show that some NTCs have consistent off-target activity,
# these will be excluded, it's the blue dots always left-low below the dashed line.
# We initially had 60 NTCs but removbed 15 because of the off-targeting that was consistent.
not_in_paper1 <-
  (elements$`voom_Hox-WT_Final_vs_Hox-WT_Input` | elements$`voom_Hox-URE_Final_vs_Hox-URE_Input`) /
    (elements$`voom_URE-AML_Final_vs_URE-AML_Input` | plot_spacer())

not_in_paper1

# Identify the names of these 15 NTCs: We reported this to the company and they said they would consider not using them anymore
luc_blacklisted <-
  lapply(c("Hox-WT_Final_vs_Hox-WT_Input", "Hox-URE_Final_vs_Hox-URE_Input", "URE-AML_Final_vs_URE-AML_Input"), function(x) {
    res_voom[[x]] %>%
      dplyr::filter(grepl("^NTC", shRNA)) %>%
      dplyr::filter(logFC < -log2(1.5)) %>%
      dplyr::pull(shRNA) %>%
      sort()
  }) %>% unlist()

# 3 means that in all three celllines these were identified as off-targets
table(luc_blacklisted)
```

# Call essential genes

```{r shrna_essential_genes_prep}

raw_counts_filtered <- raw_counts_shrna %>% filter(!rownames(.) %in% luc_blacklisted)

# Apply the trick to fool MAGeCK with the NTCs.
use_luc <-
  raw_counts_filtered %>%
  dplyr::filter(grepl("NTC", rownames(raw_counts_filtered))) %>%
  set_rownames(gsub("NTC", "NTC_for_norm", rownames(.)))

# That is the count table one would save to disk and run MAGeCK on.
counts_for_mageck <-
  rbind(
    raw_counts_filtered, # here this is the actual counts
    use_luc
  ) %>% # and this is the "copied" NTCs
  tibble::rownames_to_column("shrna") %>%
  dplyr::mutate(gene = gsub("__.*", "", shrna)) %>%
  dplyr::relocate(shrna, gene)

mageck_dir <- paste0(outdir, "/mageck/")

suppressWarnings(dir.create(mageck_dir, recursive = TRUE))

write.table(
  x = data.frame(x = grep("_for_norm", counts_for_mageck$gene, value = TRUE)),
  file = paste0(mageck_dir, "/controls_ntc_mageck.txt"),
  col.names = FALSE, row.names = FALSE, quote = FALSE, sep = "\n"
)

write.table(
  x = counts_for_mageck,
  file = paste0(mageck_dir, "/counts_for_mageck.txt"),
  col.names = TRUE, row.names = FALSE, quote = FALSE, sep = "\t"
)

Sys.setenv(MAGECKDIR = mageck_dir)
```

# Run MAGeCK

```{bash run_mageck}

# MAGeCK is available in this Docker container

# Run MAGeCK for each cell line separately. The names provided via the 
# -t and -c argument tell MAGeCK which columns of the count matrix to use during
# the current iteration. The 'controls_ntc.txt' contains the names of the rows
# of the count matrix to use as normalization / null permutation controls,
# so that is the 45 NTCs.

export PATH="$PATH:$MAGECKPATH"

cd "$MAGECKDIR"

echo 'HoxWT HoxURE B22' \
| tr " " "\n" \
| while read p; do
    mageck test \
      -k counts_for_mageck.txt \
      -t ${p}_Final_rep1,${p}_Final_rep2,${p}_Final_rep3 \
      -c ${p}_Input_rep1,${p}_Input_rep2,${p}_Input_rep3 \
      --control-gene controls_ntc_mageck.txt --norm-method control --gene-lfc-method alphamedian \
      -n ${p} --additional-rra-parameters '--min-number-goodsgrna 3 --permutation 10000' --keep-tmp \
      --normcounts-to-file
  done < /dev/stdin    
  
```

# Analyse MAGeCK results

Require that essential genes have a `neg.fdr` below 0.01. The MAGeCK call above also forced a minimum of
three "good shRNAs", meaning that three out of five shRNAs must support the call of being an essential gene.

```{r shrna_essential_gene_analysis}

# Read the MAGeCK output we provide in the supplement
WT_RRA_gene <- read.delim(paste0(mageck_dir, "/HoxWT.gene_summary.txt"))
WT_RRA_shRNA <- read.delim(paste0(mageck_dir, "/HoxWT.sgrna_summary.txt"))
URE_RRA_gene <- read.delim(paste0(mageck_dir, "/HoxURE.gene_summary.txt"))
URE_RRA_shRNA <- read.delim(paste0(mageck_dir, "/HoxURE.sgrna_summary.txt"))
B22_RRA_gene <- read.delim(paste0(mageck_dir, "/B22.gene_summary.txt"))
B22_RRA_shRNA <- read.delim(paste0(mageck_dir, "/B22.sgrna_summary.txt"))

# Killing and NTCs were defined by us during library design
killing_controls <- c("Psma1", "Rpl30", "Polr2b")
negative_control <- "NTC"

# Extract essential genes as those with FDR < 0.01
mageck_results <- c("WT_RRA_gene", "URE_RRA_gene", "B22_RRA_gene")

essential_genes_full <-
  sapply(mageck_results, function(x) {
    l <- get(x) %>% filter(!grepl(paste(c(killing_controls, negative_control), collapse = "|"), id))
    l %>% filter(neg.fdr < 0.01)
  }, simplify = FALSE)

essential_genes <- sapply(essential_genes_full, function(x) x$id, simplify = FALSE)

names(essential_genes_full) <- paste0("MAGeCK_results_", gsub("_RRA_gene", "", names(essential_genes)))

# This is essentially Dataset EV3
sapply(essential_genes_full, head, simplify = FALSE)

# Collect per-shRNA logFCs, then quantile normalize
RRA_logFC_raw <-
  Reduce(
    function(x, y) merge(x, y, all = TRUE, by = "shRNA"),
    list(
      data.frame(shRNA = WT_RRA_shRNA$sgrna, HoxWT = WT_RRA_shRNA$LFC),
      data.frame(shRNA = URE_RRA_shRNA$sgrna, HoxURE = URE_RRA_shRNA$LFC),
      data.frame(shRNA = B22_RRA_shRNA$sgrna, B22 = B22_RRA_shRNA$LFC)
    )
  ) %>%
  dplyr::mutate(Gene = gsub("__.*", "", shRNA)) %>%
  dplyr::relocate(shRNA, Gene)

RRA_logFC_qn <-
  cbind(
    RRA_logFC_raw %>% select(contains(c("shRNA", "Gene"))),
    limma::normalizeQuantiles(RRA_logFC_raw %>% dplyr::select(contains(c("Hox", "B22"))))
  )

Figure_3A <-
  RRA_logFC_qn %>%
  reshape2::melt(variable.name = "Cellline", value.name = "logFC", id.vars = c("shRNA", "Gene")) %>%
  mutate(
    label = case_when(
      Gene %in% killing_controls ~ "killing",
      grepl(negative_control, Gene) ~ "NTC",
      TRUE ~ "signature3"
    ),
    Cellline = factor(gsub("Hox", "Hox-", gsub("B22", "URE-AML", as.character(Cellline))),
      levels = c("Hox-WT", "Hox-URE", "URE-AML")
    )
  ) %>%
  ggplot(aes(x = Cellline, y = logFC, color = label, size = label)) +
  geom_point(position = position_jitter(seed = 1, width = .25), size = .75) +
  scale_color_manual(values = c("firebrick", "darkblue", "grey")) +
  xlab("") +
  scale_size_manual(values = c(2, 2, 2)) +
  theme(legend.title = element_blank(), legend.position = "top", legend.justification = "left") +
  guides(x = guide_axis(angle = 22), color = guide_legend(ncol = 1, override.aes = list(size = 3)))

Figure_3A

# Summarize overlap of essential genes in an upset
Figure_3B <-
  rbind(
    data.frame(Celltype = "Hox-WT", Gene = essential_genes$WT),
    data.frame(Celltype = "Hox-URE", Gene = essential_genes$URE),
    data.frame(Celltype = "URE-AML", Gene = essential_genes$B22)
  ) %>%
  dplyr::mutate(Celltype = factor(Celltype, levels = c("Hox-WT", "Hox-URE", "URE-AML"))) %>%
  dplyr::group_by(Gene) %>%
  dplyr::summarize(Celltype = list(Celltype)) %>%
  ggplot(aes(x = Celltype)) +
  geom_bar(fill = "grey35") +
  geom_label(stat = "count", aes(label = after_stat(count)), vjust = 1.5) +
  scale_x_upset(position = "bottom") +
  # theme_combmatrix(combmatrix.label.text=element_text(size=10)) +
  xlab("") +
  ylab("overlaps")

Figure_3B

# Get differential logFCs for Figure 3C
differential_logFC <-
  RRA_logFC_qn %>%
  dplyr::mutate(shRNA = shRNA, Gene = Gene, URE_WT = HoxURE - HoxWT, B22_WT = B22 - HoxWT) %>%
  tibble::column_to_rownames("shRNA") %>%
  dplyr::select(c(Gene, contains(c("_")))) %>%
  tibble::remove_rownames()

empirical_cutoff <-
  apply(
    differential_logFC %>%
      dplyr::filter(Gene %in% negative_control) %>%
      dplyr::select(contains("_")),
    2,
    function(x) quantile(abs(x), .95)
  ) %>%
  as.numeric() %>%
  mean()

essential_merged <- unique(unlist(essential_genes))

# This table summarizes how many shRNAs per gene support a call that a gene
# is more or less essential in the given comparison.
differential_table_candidate <- lapply(essential_merged, function(x) {
  d <- differential_logFC %>% filter(Gene == x)

  d %>%
    dplyr::mutate(
      URE_more_essential_WT = sum(URE_WT < -empirical_cutoff),
      # URE_less_essential_WT=sum(URE_WT >  empirical_cutoff),
      B22_more_essential_WT = sum(B22_WT < -empirical_cutoff),
      # B22_less_essential_WT=sum(B22_WT >  empirical_cutoff)
    ) %>%
    dplyr::select(c(Gene, contains(c("_more", "_less")))) %>%
    unique() %>%
    tibble::remove_rownames()
}) %>%
  do.call(rbind, .) %>%
  tibble::column_to_rownames("Gene")

# We use the above table to filter for the final differentially-essential genes.
# This is done with regard to the WT cell line.
n_is <- 3 # we need evidence that WT is less affected than the other group for that many shRNAs
n_no <- 1 # we require that no more than that many shRNAs have evidence for the opposite pattern:
differentially_essential <-

  list(

    # this is the genes more essential to only URE compared to WT
    kills_URE = differential_table_candidate %>%
      filter(URE_more_essential_WT >= n_is &
        B22_more_essential_WT < n_is &
        rownames(.) %in% essential_genes$URE) %>%
      rownames(),

    # more essential to only B22 compared to WT
    kills_B22 = differential_table_candidate %>%
      dplyr::filter(B22_more_essential_WT >= n_is &
        URE_more_essential_WT < n_is &
        rownames(.) %in% essential_genes$B22) %>%
      rownames(),

    # more essential to URE & B22 compared to WT
    kills_both = differential_table_candidate %>%
      dplyr::filter(B22_more_essential_WT >= n_is &
        URE_more_essential_WT >= n_is &
        rownames(.) %in% essential_genes$URE &
        rownames(.) %in% essential_genes$B22) %>%
      rownames()
  )

# Post-filter so the "more essential" URE are essential in URE and
# "more essential" AML are essential in AML
# To-Do: Formally prefilter so that only essential genes in URE/B22 go into the differential analysis
# lapply(names(differentially_essential), function(x){


# }

# This is then the input for this wonderful monolithic chunk of heatmap code.
matrix_toplot <- differential_table_candidate %>%
  dplyr::filter(rownames(.) %in%
    unlist(differentially_essential))

# Split the heatmap based on the three entries in differentially_essential
use_row_order <- sapply(differentially_essential,
  function(x) match(x, rownames(matrix_toplot)),
  simplify = FALSE
)

use_row_split <- unlist(lapply(
  names(use_row_order),
  function(x) rep(x, length(use_row_order[[x]]))
))

# Some aesthetic parameters for ComplexHeatmap
new_fs <- 4
hm_fs <- gg2gp(new_fs)
ht_opt$heatmap_row_names_gp <- grid::gpar(fontsize = gg2gp(new_fs))
ht_opt$heatmap_column_names_gp <- grid::gpar(fontsize = gg2gp(new_fs))
ht_opt$heatmap_column_title_gp <- grid::gpar(fontsize = gg2gp(new_fs))
ht_opt$heatmap_row_title_gp <- grid::gpar(fontsize = gg2gp(new_fs))

# Plot heatmap with number of shRNAs supporting the call as being differentially essential
hm_toplot <- matrix_toplot[unlist(use_row_order), ] %>% dplyr::select(dplyr::contains("more"))
rownames(hm_toplot) <- gsub(".*_", "", rownames(hm_toplot))
colnames(hm_toplot) <- gsub("^URE", "Hox-URE", gsub("WT", "Hox-WT", gsub("B22", "URE-AML", colnames(hm_toplot))))

pdf(NULL)
Figure_3C <-
  draw(Heatmap(
    matrix = as.matrix(hm_toplot),
    col = c("grey95", "grey90", "grey80", "#90A4B4FF", "#638FB4FF", "#0966B4FF"),
    row_names_gp = grid::gpar(fontsize = hm_fs),
    column_title_gp = grid::gpar(fontsize = hm_fs),
    row_title_gp = grid::gpar(fontsize = hm_fs),
    column_names_gp = grid::gpar(fontsize = hm_fs),
    heatmap_legend_param = list(
      legend_direction = "horizontal",
      legend_width = unit(4, "cm"),
      legend_position = "bottom",
      at = c(0, 1, 2, 3, 4, 5),
      title_gp = gpar(fontface = "plain", fontsize = gg2gp(new_fs)), nrow = 1
    ),
    name = "number of supporting shRNAs per gene",
    cluster_columns = FALSE,
    column_title = NULL,
    show_row_dend = FALSE,
    column_labels = gt_render(gsub("_more_essential_", " > ", colnames(hm_toplot)),
      rot = 0, padding = unit(c(20, -10, 0, 0), "pt")
    ),
    row_split = gsub("_", " ", use_row_split), row_title = NULL
  ),
  heatmap_legend_side = "top"
  )
invisible(dev.off())

Figure_3C
```

This chunk is not executed, but documents what was uploaded to the GEO submission as processed files:

```{r, eval = FALSE}

supplement_shrnascreen <- paste0(outdir, "/lists/shrnascreen/")
suppressWarnings(dir.create(supplement_shrnascreen))

out <- paste0(supplement_shrnascreen, "/counts_raw.txt")
data.table::fwrite(x = counts_for_mageck, file = out, col.names = TRUE, row.names = FALSE, sep = "\t", quote = FALSE)
system(paste0("gzip --force --best ", out))
rm(out)
```
